---
title: "Recoding GSS religion"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

rec_relig_12 <- function(religion, denomination, other)  {
  ######## Recoding religion and religion at 16 ################
  # Sherkat and Lehman (2017)
  # To work properly, folder 'religion_punches' with .csvs
  # of label names has to be in wokring directory.
  #
  # Function: # rec_relig_12(religion, denomination, other)
  #	relig or relig16 variable; denom or denom 16; other or oth16
  # function prints frequencies and returns factor vector with
  # religion recoded.
  # 
  # It works with GSS dataset imported through 'read.spss',
  # from foreign package, in following way:
  # to.data.frame = TRUE, trim.factor.names = TRUE,
  # trim_values = TRUE, use.missings = FALSE
  
  ## SETTING UP ==================================================
  
  # Import three varaibles into new dataset used for recoding ----
  DF <- data.frame(relig = religion,
                   denom = denomination,
                   other = other
                   )
  
  # Read values for all variables --------------------------------
  c_relig <- read.csv("religion_punches/relig.csv")
  c_denom <- read.csv("religion_punches/denom.csv")
  c_other <- read.csv("religion_punches/other.csv")


  # Create vectors with position corespondign to the punch
  # of label in DF codebook for 3 variables 

  c_r <- c()
  for (i in c_relig$punch)  {
    c_r[i] <- as.character(c_relig[c_relig$punch == i, "label"])
  }
  c_r[99] <- "NA"

  c_d <- c()
  for (i in c_denom$punch)  {
    c_d[i] <- as.character(c_denom[c_denom$punch == i, "label"])
  }
  c_d[99] <- "NA"
  
  c_o <- c()
  for (i in c_other$punch)  {
    c_o[i] <- as.character(c_other[c_other$punch == i, "label"])
  }
  c_o[999] <- "NA"
  
  
  ## RECODING =====================================================
  
  # Liberal Protestants --------------------------------------------
  lp_d_num <- c(40:49)
  lp_o_num <- c(29, 30, 40, 54, 70, 72 , 81, 82, 95, 98, 119,
                142, 160, 188)
  lp_denom <- c_d[lp_d_num]
  lp_other <- c_o[lp_o_num]
  DF$lp_true <- (DF$denom %in% lp_denom) | (DF$other %in% lp_other)
  DF$rv[DF$lp_true] <- "Liberal Protestant"
  
  # Episcopalians --------------------------------------------------
  ep_d_num <- c(50)
  ep_denom <- c_d[ep_d_num]
  DF$ep_true <- DF$denom %in% ep_denom
  DF$rv[DF$ep_true] <- "Episcopalian"
  
  # Moderate Protestants ------------------------------------------
  mp_d_num <- c(10:13, 20:23, 28)
  mp_o_num <- c(1, 8, 15, 19, 25, 32, 42:44, 46, 49:51, 71, 73, 94,
                99, 146, 148, 150, 186)
  mp_denom <- c_d[mp_d_num]
  mp_other <- c_o[mp_o_num]
  DF$mp_true <- (DF$denom %in% mp_denom) | (DF$other %in% mp_other)
  DF$rv[DF$mp_true] <- "Moderate Protestant"

  # Lutherans -----------------------------------------------------
  lt_d_num <- c(30:38)
  lt_o_num <- c(105)
  lt_denom <- c_d[lt_d_num]
  lt_other <- c_o[lt_o_num]
  DF$lt_true <- (DF$denom %in% lt_denom) | (DF$other %in% lt_other)
  DF$rv[DF$lt_true] <- "Lutheran"

  # Baptists ------------------------------------------------------
  bp_d_num <- c(14:18)
  bp_o_num <- c(93, 133, 197)
  bp_denom <- c_d[bp_d_num]
  bp_other <- c_o[bp_o_num]
  DF$bp_true <- (DF$denom %in% bp_denom) | (DF$other %in% bp_other)
  DF$rv[DF$bp_true] <- "Baptist"

  # Sectarian Protestants -----------------------------------------
  # these initial variables pull out sectarians codes
  # relig == 11 (christian) or relig == 5 (other),
  # but also have valid denom codes. 
  DF$sp_pent <- (DF$relig == c_r[11]) & (DF$other == c_o[68])
  DF$sp_centchrist <- (DF$relig == c_r[5]) & (DF$other == c_o[31])
  DF$sp_fsg <- (DF$relig == c_r[5]) & (DF$other == c_o[53])
  DF$sp_jw <- (DF$relig == c_r[5]) & (DF$other == c_o[58])
  DF$sp_sda <- (DF$relig == c_r[5]) & (DF$other == c_o[77])
  DF$sp_ofund <- (DF$relig == c_r[5]) & (DF$other == c_o[97])
  
  sp_o_num <- c(2, 3, 5:7, 9, 10, 12:14, 16:18, 20:24, 26, 27, 31,
                33:39, 41, 45, 47, 48, 52, 53, 55:58, 63, 65:69,
                76:79, 83:92, 96, 97, 100:104, 106:113, 115:118,
                120:122, 124, 125, 127:132, 134, 135, 137:141, 144,
                145, 151:156, 158, 159, 166:182, 184, 185, 187,
                189:191, 193, 195, 196, 198, 201, 204)
  sp_other <- c_o[sp_o_num]
  
  DF$sp_true <- ((DF$other %in% sp_other) |
                  DF$sp_pent |
                  DF$sp_centchrist |
                  DF$sp_fsg |
                  DF$sp_jw |
                  DF$sp_sda |
                  DF$sp_ofund
                 )
  DF$rv[DF$sp_true] <- "Sectarian Protestant"

  # Christian, no group identified --------------------------------
  DF$cn_christ <- (DF$relig == c_r[11]) & !DF$sp_pent
  cn_r_num <- c(13)
  cn_d_num <- c(70, 98, 99)
  cn_o_num <- c(998, 999)
  cn_relig <- c_r[cn_r_num]
  cn_denom <- c_d[cn_d_num]
  cn_other <- c_o[cn_o_num]
  DF$cn_true <- ((DF$relig %in% cn_relig) |
                 (DF$denom %in% cn_denom) |
                 (DF$other %in% cn_other) |
                  DF$cn_christ
                 )
  DF$rv[DF$cn_true] <- "Christian, no group given"

  # Mormons -------------------------------------------------------
  mr_o_num <- c(59:62, 64, 157, 162)
  mr_other <- c_o[mr_o_num]
  DF$mr_true <- DF$other %in% mr_other
  DF$rv[DF$mr_true] <- "Mormon"
  
  # Catholics or Orthodox Christians ------------------------------
  co_r_num <- c(2, 10)
  co_o_num <- c(28, 123, 126, 143, 149, 183, 194)
  co_relig <- c_r[co_r_num]
  co_other <- c_o[co_o_num]
  DF$co_true <- (DF$relig %in% co_relig) | (DF$other %in% co_other)
  DF$rv[DF$co_true] <- "Catholic or Orthodox"

  # Jews ----------------------------------------------------------
  jw_r_num <- c(3)
  jw_relig <- c_r[jw_r_num]
  DF$jw_true <- DF$relig %in% jw_relig
  DF$rv[DF$jw_true] <- "Jewish"
  
  # Other religions -----------------------------------------------
  DF$or_nonsp <- (DF$relig == c_r[5]) & !(DF$sp_pent |
                                          DF$sp_centchrist |
                                          DF$sp_fsg |
                                          DF$sp_jw |
                                          DF$sp_sda |
                                          DF$sp_ofund
                                          )
  or_r_num <- c(6:9, 12)
  or_o_num <- c(11, 74, 75, 80, 114, 136, 161, 163, 164, 192)
  or_relig <- c_r[or_r_num]
  or_other <- c_o[or_o_num]
  DF$or_true <- ((DF$relig %in% or_relig) |
                 (DF$other %in% or_other) |
                  DF$or_nonsp
                 )
  DF$rv[DF$or_true] <- "Other religion"
  
  # No religious identification -----------------------------------
  nr_r_num <- c(4)
  nr_relig <- c_r[nr_r_num]
  DF$nr_true <- DF$relig %in% nr_relig
  DF$rv[DF$nr_true] <- "None"
  
  # Missing values ------------------------------------------------
  # No Answer
  DF$na_relig <- DF$relig == c_r[99]
  DF$na_denom <- DF$denom == c_d[99] 
  DF$na_rd <- DF$na_relig & DF$na_denom
  DF$rv[DF$na_rd] <- "No answer"

  # Don't know
  DF$dk_relig <- DF$relig == c_r[98]
  DF$rv[DF$dk_relig] <- "Dont know"
  
  
  ## FINISHING IT =================================================

  # Treat it as a factor
  DF$rv <- as.factor(DF$rv)

  # Provide table with proportions
  print(cbind(Freq = table(DF$rv, useNA = "ifany"),
              Relative = round(100 * 
                                 prop.table(
                                   table(
                                     DF$rv,
                                     useNA = "ifany"
                                     )
                                   ),
                               2),
              Cumul = round(100 *
                              cumsum(
                                prop.table(
                                  table(
                                    DF$rv,
                                    useNA = "ifany"
                                    )
                                  )
                                ),
                            2)
              )
        )
  
  # Return the vector with recoded religion
  return(DF$rv)
}


```

***

R function that recodes GSS religion from three variables (`relig`, `denom`, `other`) to 12 categories, according to Darren E. Sherkat and Derek Lehman, ["After The Resurrection: The Field of the Sociology of Religion in the United States"](https://iranianredneck.wordpress.com/2016/11/29/why-reltrad-sucks-contesting-the-measure-of-american-religion/){target="_blank"}.^[Paper contains SPSS syntax based on which I wrote function - also some of the comments in the code are pasted from there.]

***

Main goal is to address four problems:

1. Recode religion successfully.
2. Easily (automatically) recode whole family of religion variables in GSS (religion, religion at 16, parents' religion, and special modules on friends' religion).
3. Recode no matter how GSS data is imported to R.
4. Provide two ways of grouping religion (12 and 7 groups).

At this point, only 1 works well, while 2 and 3 are partial. Beside huge number of categories (cca 200 for other), main problem is that punch codes from GSS codebook do not correspond to factor values once data is imported (some codes and blocks of numbers are just skipped).

## Example of use

```{r, message=FALSE, warning=FALSE}
library(foreign)

GSS <- read.spss("GSS2014.sav",
                 to.data.frame = TRUE,
                 trim.factor.names = TRUE,
                 trim_values = TRUE,
                 use.missings = FALSE
                 )

summary(GSS[, c("relig", "denom", "other")])


```

If we simply apply function to the GSS 2014 data,^[Data is not included to github repository since it is larger than 10MB. It can be downloaded freely from NORC website.] we get following:

```{r}
GSS$religion <- rec_relig_12(GSS$relig, GSS$denom, GSS$other)
summary(GSS$religion)
tail(GSS[, c("relig", "denom", "other", "religion")])
```

## Attention!

Function is usable but with certain limitations. At this moment it successfully works only for respondent's religion and religion at the age of 16 (although the process can be used on other derived variables), and when data is imported from SPSS file to data frame, with trimmed factor names and values, without use of missing values.

*If this does not suit your need, ad hoc workaround is to import GSS dataset in your fashion of choice (e.g. with use of `NA`) and identical dataset as required by this function. You should imediatlly use the function on the second dataframe, and assing values of new variable to the first. Ideally, you would also assing the variables used for recoding, and check them against main dataframe that everything is OK.*

## Explanation of how it works

There is probably an easier way to do this, but I started using imported GSS data from .sav files with categorical variables as factors using foreign package. However, R's nominal values of factors do not correspond to SPSS codes for factors (i.e. corresponding punches in GSS codebook) which causes problem in trying to directly replicate SPSS syntax.

Here we can see the sample from 2014 GSS file (including new recoded var):

```{r}
str(GSS[, c("relig", "denom", "other", "religion")])
```


As a starting point for a more universal recoding function, this one uses three tables (in `religion_punches` folder, heads presented here) from which three vectors are formed (`relig`, `denom`, and `other`) where index corresponds to punch number, and label is value.

```{r, message=FALSE, warning=FALSE, echo = FALSE, caption = "Heads of csv files showing problem with punches."}
relig_punch <- read.csv("religion_punches/relig.csv")
denom_punch <- read.csv("religion_punches/denom.csv")
other_punch <- read.csv("religion_punches/other.csv")

list(relig.csv = head(relig_punch[,2:3]),
     denom.csv = head(denom_punch[,2:3]),
     punch.csv = head(other_punch[,2:3])
     )
```

For each *new* religious identification, separate vectors (for `relig`, `denom`, and/or `other`) are created that contain punch codes and labels, per paper and SPSS syntax. In addition, logical vectors are created for Sectarian Protestants with valid denomination codes for sorting between 'Sectarian Protestants', 'Christian - no group given', and 'other religions'.

*This approach is somewhat superior to usual practices of recoding, since it can be confirmed through basic set operations that there is no overlap between groups or that some categories are left out.*

Next, twelve logical vectors, for respondent's belonging to each group, are created by checking against appropriate vectors containing labels. Names are assigned, function prints frequency table, and returns vector that can be assigned to a new variable.

This allows that basic structure can be generalized for recoding of all three-variable sets of religious identification that can be imported as either factors with names or numerical values (or combination, as is the case with some groups), easy recoding into different number of groups, and printing of detailed classification that includes names of all religions and denominations.


## Final remarks

Function still needs some fixing/generalization for handling different coding of variables. I will periodically update it, as I occasionally improve it.

Please feel free to contact me with any questions or comments.

If you find function handy and use it, cite the source paper; in addition acknowledging my contribution is appreciated.
